Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-20 paper/ms.tex
update ms

In the context of Kepler, rotation-based ages have been reported for
planet-hosting stars since the early data releases
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
focused on rotation-based ages for Kepler have further explored
incorporating information from stellar kinematics
\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
analysis is motivated by a few factors that we believe constitute
improvements, particularly for the ages of sub-gigayear stars.   These
factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    For Sun-like stars, this model is consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
    of uncertainty propagation.  For late G dwarfs and
    K dwarfs, the calibration differs by factors of a few;
  %
  \item We can identify binary stars at greater fidelity today than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,2022AJ....164...88B}.
    These clusters offer a sample of stars for which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

Using our effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is calibrated out
to 4\,Gyr, the age of M67
\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the
stellar spin-down rate as a function of time and stellar temperature,
the framework interpolates between reference open clusters using cubic
Hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citet{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

@@

The stellar ages and planetary properties of the systems discussed in
the previous sections are given in Tables~\ref{tab:planets}
and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
age distributions of all the stars (left column) and the known
planet-hosts (middle column).  These ``histograms'' are sampled from
the posterior probability distributions by drawing ten random samples
from each posterior distribution, and then computing the re-weighted
histogram of the resulting samples.  The uncertainties assume Poisson
statistics.  We truncated the plots near $4$\,Gyr since this is the
cluster age out to which our interpolation-based method is calibrated.
We refer to the overall slope as a ``demographic cliff'' in reference
to an impending human phenomenon forecasted to happen in coming years.

The age distributions of Kepler stars and planet hosts are
statistically quite similar.  A simple diagnostic of whether the two
distributions are drawn from the same underlying distribution is the
Kolmogorov-Smirnov test;  we calculated this statistic in a manner
that accounts for the Poisson uncertainties by performing 100 random
draws of 70\% of the stars from each sample.  The resulting median
$p$-values were $\approx$0.01 for both rows of
Figure~\ref{fig:hist_tgyro}.  This degree of similarity suggests that
the ``common wisdom'' that young transiting planets are hard to detect
because their host stars are photometrically variable requires some
nuance.  If true, this statement only holds for the very youngest
stars, since at the age resolution of this study, there appears to be
little to no selection bias against sub-gigayear planets.

@@

{\it MELANGE-3}---\citet{2022AJ....164...88B} reported a
$105\pm10$\,Myr association in the Kepler field and derived its age
using rotation, lithium, and the isochronal pre-main-sequence.
Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
this association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.

>>>>

In the context of Kepler, rotation-based ages have been reported for
planet-hosting stars since the early data releases
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
focused on rotation-based ages for Kepler have further explored
incorporating information from stellar kinematics
\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
analysis is motivated by a few factors that we believe constitute
improvements, particularly for the ages of sub-gigayear stars.   These
factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    For Sun-like stars, this model is consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
    of uncertainty propagation.  For late G dwarfs and
    K dwarfs, the calibration differs by factors of a few;
  %
  \item We can identify binary stars at greater fidelity today than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars for which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

Using our effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is calibrated out
to 4\,Gyr, the age of M67
\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the
stellar spin-down rate as a function of time and stellar temperature,
the framework interpolates between reference open clusters using cubic
Hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citet{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

@@

The stellar ages and planetary properties of the systems discussed in
the previous sections are given in Tables~\ref{tab:planets}
and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
age distributions of all the stars (left column) and the known
planet-hosts (middle column).  These ``histograms'' are sampled from
the posterior probability distributions by drawing ten random samples
from each posterior, and then computing the re-weighted
histogram of the resulting samples.  The uncertainties assume Poisson
statistics.  We truncated the plots near $4$\,Gyr since this is the
cluster age out to which our interpolation-based method is calibrated.
We refer to the overall slope as a ``demographic cliff'' in reference
to an impending human phenomenon forecasted to happen in coming years.

The age distributions of Kepler stars and planet hosts are
nearly indistinguishable.  A simple diagnostic of whether the two
distributions are drawn from the same underlying distribution is the
Kolmogorov-Smirnov test;  we calculated this statistic in a manner
that accounts for the Poisson uncertainties by performing 100 random
draws of 70\% of the stars from each sample.  The resulting median
$p$-values were $\approx$0.01 for both rows of
Figure~\ref{fig:hist_tgyro}.  This degree of similarity suggests that
the ``common wisdom'' that young transiting planets are hard to detect
because their host stars are photometrically variable requires some
nuance.  If true, this statement only holds for the very youngest
stars, since at the age resolution of this study, there appears to be
little to no selection bias against sub-gigayear planets.

@@

{\it MELANGE-3}---\citet{Barber_2022} reported a
$105\pm10$\,Myr association in the Kepler field and derived its age
using rotation, lithium, and the isochronal pre-main-sequence.
Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
this association.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{Barber_2022}.
